---
title: "Day 15"
author: "Taylor R. Brown"
date: "1/3/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Roadmap

9.1

- back to non-time series situation

- **importance sampling** -> self-normalized importance sampling

- mostly proof-free (classical asymptotics)

9.2

- sampling importance resampling

9.3

- sequential importance sampling with resampling (single-step analysis)

9.4

- sequential importance sampling with resampling (multi-step analysis)


<!-- ## Assumption 9.1.1 -->

<!-- The target distribution $\mu$ is absolutely continuous with respect to the instrumental/proposal distribution $\nu$, written $\mu \ll \nu$, and $d\mu /d\nu > 0$ ($\nu$-almost surely). -->


## Theorem 9.1.2 (first part)

Let $f$ be a real, measurable function such that $\mu(|f|) < \infty$ and $|f| \mu \ll \nu$ (note the typo in the book), and let $\xi^1, \xi^2, \ldots$ be a sequence of iid random variables from $\nu$. Then the unnormalized importance sampling estimator 
$$
\tilde{\mu}^{\text{IS}}_{\nu, N}(f) = \frac{1}{N} \sum_{i=1}^N  f(\xi^i) \frac{d\mu}{d\nu}(\xi^i)
$$
is strongly consistent. 

This is just the ordinary strong law of large numbers.


## Theorem 9.1.2 (second part)


If in addition
$$
E_{\nu}\left[f^2(\xi^i) \frac{d\mu}{d\nu}(\xi^i)^2 \right] < \infty,
$$

then $\tilde{\mu}^{\text{IS}}_{\nu, N}$ is asymptotically normal, i.e. 
$$
\sqrt{N}\left(\tilde{\mu}^{\text{IS}}_{\nu, N}(f) - \mu(f) \right) \overset{\text{D}}{\to} \text{Normal} \left(0, \text{Var}_{\nu}\left[ f(\xi^i) \frac{d\mu}{d\nu}(\xi^i)\right] \right)
$$

This is just the regular central limit theorem.

## Theorem 9.1.2 in practice

- $|f| \mu \ll \nu$ is weaker than $\mu \ll \nu$

- infinite variance estimators are quite bad in practice

- pick $\nu$ carefully! (e.g. $P_{\mu}(\xi > 1e4) = E_{\mu}[\mathbb{1}(\xi > 1e4)]$)

- a rearrangement,
\begin{align*}
\text{Var}_{\nu}\left[ f(\xi^i) \frac{d\mu}{d\nu}(\xi^i)\right] 
&= E_{\nu}\left[ \left( f(\xi^i) \frac{d\mu}{d\nu}(\xi^i) - \nu\left[ f(\xi^i) \frac{d\mu}{d\nu}(\xi^i)\right]\right)^2 \right] \\
&= E_{\nu}\left[ \left( f(\xi^i) \frac{d\mu}{d\nu}(\xi^i) - \mu(f)\right)^2 \right] \\
&= \mu^2(f)  \nu\left[ \left( \frac{f d\mu/d\nu}{\mu(f)} - 1\right)^2 \right] \\
&\ge \mu^2(f)  \nu\left[ \left( \frac{|f| d\mu/d\nu}{\mu(|f|)} - 1\right)^2 \right] \tag{typo in book}
\end{align*}

## Theorem 9.1.2 in practice


$$
\text{Var}_{\nu}\left[ f(\xi^i) \frac{d\mu}{d\nu}(\xi^i)\right] 
= \mu^2(f) \nu\left[ \left( \frac{f d\mu/d\nu}{\mu(f)} - 1\right)^2 \right]
$$


- $\frac{f d\mu/d\nu}{\mu(f)}$ has mean $1$ no matter what $\nu$ is, so we want to minimize the second moment or variance.

- $\frac{f d\mu/d\nu}{\mu(f)}$ close to $1$ with low variance means $f d\mu/d\nu$ is close-to-constant or flat. 

- if $f$ is negligible, this means $\mu$ and $\nu$ are roughly the same. 

- so, this is really only more useful than regular Monte-Carlo if $f$ is not negligible (which is pretty much always true).


## Example 9.1.3

- target: $\mu$ is $\text{Cauchy}(0,1)$ (fat tails)

- proposal: $\nu$ is $\text{Normal}(0,1)$ (thin tails)

$$
\frac{d\mu}{d\nu}(x) = \frac{\frac{1}{\pi}\left(1 + x^2\right)^{-1} }{\frac{1}{\sqrt{2\pi}}\exp[-.5x^2] } \propto \exp[.5x^2]/(1+x^2)
$$

## Example 9.1.3

plotting the RN derivative...
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
plot(x_grid, dt(x_grid, df=1)/dnorm(x_grid), type="l") # not flat!
```


## Example 9.1.3

plotting $x \frac{d\mu}{d\nu}(x)$
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
plot(x_grid, x_grid * dt(x_grid, df=1)/dnorm(x_grid), type="l") # not flat!
```


## Example 9.1.3

plotting $\mathbf{1}(|x| < 3) \frac{d\mu}{d\nu}(x)$
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
silly_indicator <- function(x) ifelse(abs(x) < 3, 1, 0)
plot(x_grid, silly_indicator(x_grid) * dt(x_grid, df=1)/dnorm(x_grid), type="l") # not too un-flat
```

## Example 9.1.4

now switch the target and the proposal!

- proposal: $\mu$ is $\text{Cauchy}(0,1)$ (fat tails)

- target: $\nu$ is $\text{Normal}(0,1)$ (thin tails)

$$
\frac{d\mu}{d\nu}(x) \propto (1+x^2)/\exp[.5x^2]
$$

## Example 9.1.4

plotting the RN derivative...
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
plot(x_grid, dnorm(x_grid)/dt(x_grid, df=1), type="l") # better!
```


## Example 9.1.4

plotting $x \frac{d\mu}{d\nu}(x)$
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
plot(x_grid, x_grid * dnorm(x_grid)/dt(x_grid, df=1), type="l") # better!
```

## Example 9.1.4

plotting $\mathbf{1}(|x| < 3) \frac{d\mu}{d\nu}(x)$
```{r, echo=TRUE}
x_grid <- seq(-10,10,.01)
plot(x_grid, silly_indicator(x_grid) * dnorm(x_grid)/dt(x_grid, df=1), type="l") # better!
```